National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions

### 2024-04-12 analysis/paper/paper.Rmd
Writing ongoing

Across 58 study areas for threatened boreal caribou (_Rangifer tarandus caribou_) in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed, and no populations of caribou are considered self-sustaining in study areas with high anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020]. The causes of this relationship include direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022;@eccc_scientific_2011]. Decreases in recruitment as disturbance increases are also observed over time in places where there is sufficient historic demographic data [@rudolph_demographic_2017;@hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships  [@johnson_science_2020]. There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020;@neufeld_primary_2021; @fortin_forest_2017; @gagne_impact_2016], and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships [@stewart_climate-informed_2023;@dyson_existing_2022].
Boreal caribou are widely distributed across the Canadian boreal region [@eccc_scientific_2011], inhabiting some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022;@ibisch_global_2016;@venter_sixteen_2016]. Monitoring rare and widely distributed species such as boreal caribou is expensive, and resources are limited.  Analysis of the potential implications of monitoring outcomes for decisions can clarify that additional monitoring is not always necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012]. A recent analysis of action plans for Canadian species at risk found half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022]. For populations of caribou and other species at risk, deciding whether or not management actions will cause declines is important.  Ecological simulation models can be used to evaluate the ability of alternative monitoring options to reduce the risk of errors in species status classifications and population viability projections [@dunham_strategic_2023].

Efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. In Bayesian population models, all available demographic or population data (e.g. abundance surveys, vital rate estimates, trend data) are combined into a single analysis that can permit novel parameter estimation, and improve precision, and account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021]. These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021]. However Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships, ignoring an important source of information about outcomes for boreal caribou across their range that can help reduce uncertainty in local demographic projections and status assessments, and limiting their ability to project outcomes on changing landscapes. Existing models have also ignored uncertainty associated with misclassification of adults and missing calves in composition surveys [@ellington_divergent_2020; @decesare_estimating_2012]. 
Here, we present a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data and accounts for composition survey errors, building on the work of Eacker et al. (2019). Following the framework proposed by Dunham et al. (2023), we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023;@dyson_existing_2022] to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. We also offer an open source app that allows users to analyze existing local demographic data, or to explore alternative monitoring scenarios with simulated data.

@@

## Caribou population monitoring: survival and recruitment metrics

Caribou demographic data are often obtained by tracking cows with telemetry collars [e.g. @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@johnson_science_2020]. Although genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g. @hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021;@flasko_fecal_2017;@jones_novel_2023; @mcfarlane_identification_2022; @mcfarlane_spatial_2021], there is still heavy reliance on collaring. Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou).

To estimate recruitment, collared females are often located and surveyed by aircraft to estimate the ratio of calves to cows in the groups of animals that include collared animals [@nbckc_nbckc_2022]. Groups typically contain uncollared animals so sample size is typically larger for recruitment than for survival estimates (see multiplier parameter $w$, cowMult below). However, a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020]. Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa (parameters $\dot{u}$ and $u$ below), and failing to detect calves (parameters $\dot{z}$ and $z$ below) [@ellington_divergent_2020; @decesare_estimating_2012]. We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details). Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020].

@@

```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios. Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%). These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, not to reflect real landscape change. In all scenarios, the projection period begins in 2024.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay3aSf1.png"))

@@

We focus on exploring a set of scenarios in which
anthropogenic disturbance, measured as the percentage of area within 500
m of disturbance [@johnson_science_2020; @eccc_scientific_2011],
increases by 1% per year (Figure \@ref(fig:figDisturbance)). The duration of monitoring varies among scenarios from 1 to 24 years. The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years. The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through med-low (20\%) and med-high (40\%) to high (60\%). These simple hypothetical scenarios span a large range to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change. Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.

@@

To simulate demographic trajectories for example populations
we used a modified version of Johnson et al's [-@johnson_science_2020]
two-stage demographic model described by Dyson et al.
[-@dyson_existing_2022]. Given a population of post-juvenile females at the beginning of year $t$, $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is binomially
distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification
biases (Supplement B) and (optionally) a delayed reproduction term to account for the reality that yearlings generally do not breed
[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum potential recruitment of female calves per cow $\dot{X}_t$:
\begin{equation}
  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
  (\#eq:dotX)
\end{equation}
where $w$ is a multiplier that defines the apparent number of adult
females in the composition survey as a function of the number of
collared animals, $\dot{q}$ is the ratio of young bulls to adult females, $\dot{u}$
is the probability of misidentifying young bulls as adult females and
vice versa, and $\dot{z}$ is the probability of missing a calf. We examine sensitivity to these unknown parameters by selecting values for each simulated example population from plausible ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and $\dot{q}$ varies uniformly between 0 and 0.6. Other authors have noted these sources of error [@ellington_divergent_2020; @decesare_estimating_2012], but our method of including uncertainty associated with unknown probabilities of missing calves and misclassification in recruitment surveys is novel.

@@

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction. White dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}

@@

Where appropriate, the Bayesian population model and national demographic model share the same structure, and we have selected notation to highlight the similarities. The difference between equations \@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to simulate observations. In Bayesian model observations are provided, and used to estimate the unobserved recruitment rate $R_t$.

@@

Following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the \ital{survival} R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a Gaussian
distributed function of the true survival probability:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).
  (\#eq:hatSEst)
\end{equation}

@@

<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives
biased results in years when \<12 months of survival data is available.
\* Note I am dropping survival information for animals that are not
alive at the beginning of a year to deal with censoring issues. There is
probably some more elegant way to deal with this, but it won't make any
difference as long as the collar deployment month is set to 1. \*
consider & comment on alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
example in all particulars, change some aspects, or do comparisons? \*
-->
We use the open source Gibbs sampling software JAGS [@plummer_jags_2023] and the \ital{R2jags} R package [@su_r2jags_2021] to fit the Bayesian model. We ran 15,000 iterations of 4 chain, discarded the first 10,000 interations, and used a thinning rate of 2. We assessed convergence for a small random subset of example trajectories using a threshold for the Gelman-Rubin diagnostic [@gelman_inference_1992] of r-hat < 1.1, and visually inspected posterior autocorrelation and traceplots produced by the \ital{mcmcplots} R package [@curtis_mcmcplots_2018].

@@

Following Dunham et al. [-@dunham_strategic_2023] we also calculate the expected value of sample information (EVSI) and the expected value of perfect information (EVPI) [@runge_which_2011;@yokota_value_2004], assigning a value of 1 to correct status assessments, and a value of 0 to incorrect assessments ($\nu_{\alpha,\psi}=1 \text{ when } \alpha=\psi, 0 \text{ otherwise}$). The expected value with perfect knowledge is $EV_{certainty}=1$. The difference between this and the expected score using knowledge of disturbance the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and priors from the national model gives EVPI, a measure of the maximum possible benefit that can be gained from local sampling:
\begin{equation}
EVPI_{\tau} = 1-max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
  (\#eq:EVPI)
\end{equation}
$P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior. EVSI measures expected improvement from a particular sampling strategy as the difference between the expected value with and without additional sample information: 
\begin{equation}
EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - (1-EVPI_{\tau}),
  (\#eq:EVSI)
\end{equation}
where $x$ is the sample information from a particular example trajectory. Simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) represent the distribution of possible values of $x$. $P(\psi|x,\tau)$ is the posterior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the sample information.

@@

Analysis was implemented using the R language [@r_core_team_r_2023] and a variety of \ital{tidyverse} packages [@wickham_welcome_2019]. To make the work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios, borrowing and building on code from an app developed by Eacker et al. [-@eacker_web-based_2019] where appropriate. Our app allows users to provide existing local data, or to make projections informed by simulated data as we have done in this analysis (similar to Figure
\@ref(fig:examplePredictions)). The app allows advanced users to adjust many model parameters (supplemental Table A1). Default parameter values are set to allow novice users to get reasonable results quickly. Most of the code needed to run the app is in an R package, \ital{caribouMetrics}, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g. @dyson_existing_2022; @stewart_climate-informed_2023]. The \ital{caribouMetrics} package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>. Find documentation and vignettes on the package website
<https://landscitech.github.io/caribouMetrics/>. The Shiny user interface app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>. Code required to run analyses and reproduce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.

@@

Analysis of simulated example demographic trajectories confirms that our Bayesian model can effectively combine local demographic data with priors from national demographic-disturbance relationships to reduce uncertainty in boreal caribou demographic projections (Figs \@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected, uncertainty about the population growth rate decreases as sampling effort in each year (i.e. number of collars) or the number of years of monitoring (Fig \@ref(fig:figDiffs)) increase. Uncertainty is higher when the forecast horizon is long (compare 5 and 20 year projections in Fig \@ref(fig:figDiffs)). Uncertainty is also higher in our high anthropogenic disturbance scenarios (Fig \@ref(fig:figDiffs)) because stochastic variation in population growth rates increase as populations become very small (as in top right panel of Fig \@ref(fig:examplePredictions)). These results confirm the Bayesian population model is working as it should, and none of them are surprising.

>>>>

The number of species threatened with extinction is high and getting higher [@ipbes_global_2019]. Reliable information on the status of populations and likely impacts of management are needed to inform conservation action but effective monitoring is challenging and costly [@legg_why_2006; @lindenmayer_effective_2018;@wintle_allocating_2010;@dunham_strategic_2023]. Resources to support conservation and recovery are limited [@buxton_balancing_2022;@mccarthy_financial_2012;@gerber_conservation_2016;@wintle_spending_2019], delays increase risk [@martin_acting_2012], and some populations have been monitored to extinction [@lindenmayer_counting_2013]. A recent analysis of action plans for Canadian species at risk found half of tasks are research and monitoring, and this proportion is highest for species with higher risk of extinction, highlighting a need to more carefully consider when, where, and why additional information is needed to inform conservation action [@buxton_balancing_2022]. Analysis of the implications of monitoring outcomes for decisions can help clarify whether additional monitoring is necessary to inform conservation action [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012].

Analysis methods that can help make better use of available monitoring data to inform decisions include Bayesian population modelling [@schaub_integrated_2021] and value of information analyses [@bennett_when_2018; @dunham_strategic_2023;@runge_which_2011; @gregory_structured_2012]. To be useful, these methods must be tailored to particular species and decision contexts. Here, we offer a Bayesian population model for boreal caribou (_Rangifer tarandus caribou_) that integrates information about the impacts of landscape change with available demographic data to reduce uncertainty in population viability projections for boreal caribou [building on @eacker_web-based_2019].  Following the value of information analysis framework proposed by Dunham et al. (2023), we combine this Bayesian model with simulations of plausible population trajectories [@stewart_climate-informed_2023;@dyson_existing_2022] to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. To help make the approach more accessible, we offer an open source app that allows users to analyze existing local demographic data, or to explore alternative monitoring scenarios with simulated data. Our approach is tailored to boreal caribou, but could be adapted for other species at risk found in multiple populations that are costly to monitor and sensitive to habitat disturbance.

Boreal caribou are widely and sparsely distributed through the Canadian boreal region across a gradient of anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020] that includes some of the most intact and inaccessible places remaining on earth [@hirsh-pearson_canadas_2022;@ibisch_global_2016;@venter_sixteen_2016]. Across 58 boreal caribou study areas, both survival and recruitment decrease with the percentage of the study area that is disturbed, and no populations of caribou are considered self-sustaining in study areas with high anthropogenic disturbance [@eccc_scientific_2011; @johnson_science_2020]. The causes of this relationship include direct impacts on habitat availability and use, and indirect impacts on predators and alternate prey [@nbckc_boreal_2022;@eccc_scientific_2011]. Decreases in recruitment as disturbance increases are also observed over time in places where there is sufficient historic demographic data [@rudolph_demographic_2017;@hervieux_widespread_2013], but many places lack sufficient data to characterize local demographic-disturbance relationships  [@johnson_science_2020]. There is also variation in demographic rates among study areas, particularly where anthropogenic disturbance is low [@johnson_science_2020;@neufeld_primary_2021; @fortin_forest_2017; @gagne_impact_2016], and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships [@stewart_climate-informed_2023;@dyson_existing_2022].

The most widely available source of boreal caribou demographic information is from cows fitted with telemetry collars [e.g. @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@decesare_estimating_2012], though genetic methods are playing an increasingly important role in some caribou monitoring programs [e.g. @hettinga_estimating_2012; @mcfarlane_genetic_2018; @moeller_integrating_2021;@flasko_fecal_2017;@jones_novel_2023; @mcfarlane_identification_2022; @mcfarlane_spatial_2021]. Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with with collars that send a mortality signal when no longer active [see @nbckc_nbckc_2021], thus allowing estimation of adult survival (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou). Recruitment is estimated from aerial surveys of the composition of groups of animals that include collared animals [@nbckc_nbckc_2022]. Groups typically also contain uncollared animals, and a critical assumption in these surveys is that different demographic groups are equally detectable [@ellington_divergent_2020]. Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa , and failing to detect calves [@ellington_divergent_2020; @decesare_estimating_2012].

Survival and recruitment survey data can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for many herd viability assessments [@decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019; @eccc_scientific_2011; @johnson_science_2020]. Some efforts have been made to improve caribou population viability projections and growth rate estimates using Bayesian methods [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019], following broader trends in the analysis of population dynamics [@schaub_integrated_2021]. In Bayesian population models, all available demographic or population data (e.g. abundance surveys, vital rate estimates, trend data) are combined into a single analysis that can permit novel parameter estimation, and improve precision, and account for multiple sources of uncertainty [@schaub_integrated_2011; @eacker_assessing_2017;@schaub_integrated_2021]. These models are proving broadly useful for rapid detection of population declines, evaluation of management actions, and viability analysis of critically endangered species [@schaub_integrated_2021]. However Bayesian caribou population models [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019] have not thus far included demographic-disturbance relationships or composition survey errors.

Here, we present a Bayesian population model for boreal caribou [building on @eacker_web-based_2019] designed to reduce uncertainty in projections of the impacts of changing landscape conditions by integrating prior information from national analysis of demographic-disturbance relationships with available local demographic data. Information is shared among populations using priors instead of integrating multiple populations into a hierarchical analysis [@leasure_hierarchical_2019] because obtaining permission to use caribou demographic data can be difficult, and most managers only have access to their own data. This Bayesian approach represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability [@eccc_scientific_2011;@johnson_science_2020] because it reconciles and accounts for uncertainties in local demographic data and demographic-disturbance relationships. We show that integrating the most informative predictor of outcomes for boreal caribou across their range can substantially reduce uncertainty in local demographic projections and status assessments, improve viability projections on changing landscapes, and help clarify where, why and when additional data is required to inform decisions and project outcomes on changing landscapes. In cases where available information is sufficient for assessing and projecting population viability, there is a risk of misallocating limited resources to monitoring that could be better spent on conservation action.

While we believe that using available information to reduce uncertainty is important, we also believe it is important to acknowledge remaining uncertainty. In general, parametric Bayesian population models include a number of simplifying assumptions, and misspecified models can yield biased or overconfident results [@schaub_integrated_2021]. More specifically, models that do not account for observation errors can be misleading [@schaub_integrated_2021]. Several authors have pointed out that  misclassification of adults and missing calves in composition surveys can bias recruitment estimates from those surveys [@ellington_divergent_2020; @decesare_estimating_2012], but none have thus far examined sensitivity of Bayesian caribou population models to these sources of error [@moeller_integrating_2021;@mcnay_demographic_2022;@eacker_web-based_2019]. We address this gap with a new bias correction term derived from a simple model of the recruitment survey observation process.

@@

## Caribou population monitoring data and composition survey bias correction

We focus on analyzing survival of collared females (as in @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019), and the ratio of calves to cows in aerial surveys of groups that include collared females [@nbckc_nbckc_2022], because these remain the most widely available sources of demographic information for boreal caribou populations [@johnson_science_2020]. Surveyed groups also contain uncollared animals so sample size is typically larger for recruitment than for survival estimates. Errors that can bias composition survey results include misidentifying young bulls as cows or vice versa, and failing to detect calves [@ellington_divergent_2020; @decesare_estimating_2012].

To derive a composition survey bias correction term we assume each group of animals in a calf:cow composition survey contains one or more collared adult females ($T$), and may also include: uncollared adult females misidentified as young bulls or unknown sex ($U$); correctly identified uncollared adult females ($V$); young bulls correctly identified as male or unknown sex ($O$); young bulls misidentified as uncollared adult females ($P$); observed calves ($J$); and unobserved calves ($K$). The apparent number of adult females in the group is $T+V+P=Tw$, where $w$ is a multiplier that defines the apparent number of adult females as a function of the number of collared animals. The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and $P=(O+P)u$. Given a probability $z$ of missing calves, we get $J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and bias-corrected recruitment rate $X$ can be written as function of the observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young bulls to adult females $q$, and the misidentification probabilities $u$ and $z$ (see Supplement B for details). For simplicity, we combine the probabilities into a bias correction term $c$ that is integrated into the Bayesian population model and simulations (below): 
$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$

@@

```{r figDisturbance, fig.cap="Anthropogenic disturbance and monitoring scenarios. Anthropogenic disturbance increases by 1% per year, and the amount of disturbance at the beginning of the projection period varies from low (0%), through low-med (20%) and med-high (40%) to high (60%). These simple hypothetical scenarios are intended to show how the need for local monitoring to reduce uncertainty varies with landscape state, and not intended to reflect real landscape change. In all scenarios, the projection period begins in 2024, and the monitoring period occurs prior to the projection period.", cache=TRUE}
knitr::include_graphics(here::here("figs/forPaper/distScnscmult6ay3aSf1.png"))

@@

We focus on exploring a set of scenarios in which
anthropogenic disturbance, measured as the percentage of area within 500
m of disturbance [@johnson_science_2020; @eccc_scientific_2011],
increases by 1% per year (Figure \@ref(fig:figDisturbance)). The duration of monitoring prior to the projection period varies among scenarios from 1 to 24 years. The projection period begins in 2024, and we consider how uncertainty changes as the projection period increases from 5 to 20 years. The amount of anthropogenic disturbance at the beginning of the projection period varies from low (0\%) through med-low (20\%) and med-high (40\%) to high (60\%). These simple hypothetical scenarios span a large range to show how the need for local monitoring to reduce uncertainty depends on landscape state, not to represent the reality of landscape change. Scenarios in which anthropogenic disturbance remains the same or decreases over time are included in Supplement D.

@@

To simulate demographic trajectories for example populations
we used a modified version of Johnson et al's [-@johnson_science_2020]
two-stage demographic model described by Dyson et al.
[-@dyson_existing_2022]. Given a population of post-juvenile females at the beginning of year $t$, $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is binomially
distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. The calf:cow ratio $\dot{R}_t$ is adjusted for sex ratio, misidentification
biases (Supplement B) and (optionally) a delayed reproduction term to account for the reality that yearlings generally do not breed
[@eacker_web-based_2019; @decesare_estimating_2012] to get maximum potential recruitment of female calves per cow $\dot{X}_t$:
\begin{equation}
  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
  (\#eq:dotX)
\end{equation}
where $w$ is a multiplier that defines the apparent number of adult
females in the composition survey as a function of the number of
collared animals, $\dot{q}$ is the ratio of young bulls to adult females, $\dot{u}$
is the probability of misidentifying young bulls as adult females and
vice versa, and $\dot{z}$ is the probability of missing a calf. We examine sensitivity to these unknown parameters by selecting values for each simulated example population from plausible ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and $\dot{q}$ varies uniformly between 0 and 0.6.

@@

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships with (top row) and without (bottom row) delayed reproduction. Orange dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models without adjustment for misidentification bias and delayed reproduction.", cache=TRUE}

@@

Where appropriate, the Bayesian population model and national demographic model share the same structure, and we have selected notation to highlight the similarities. The difference between equations \@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to simulate observations. In the Bayesian model, in contrast, observations are provided, and used to estimate the unobserved recruitment rate $R_t$.

@@

Following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the _survival_ R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a Gaussian
distributed function of the true survival probability:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).
  (\#eq:hatSEst)
\end{equation}

@@

<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives
biased results in years when \<12 months of survival data is available.
\* Note I am dropping survival information for animals that are not
alive at the beginning of a year to deal with censoring issues. There is
probably some more elegant way to deal with this, but it won't make any
difference as long as the collar deployment month is set to 1. \*
consider & comment on alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
example in all particulars, change some aspects, or do comparisons? \*
-->
We use the open source Gibbs sampling software JAGS [@plummer_jags_2023] and the _R2jags_ R package [@su_r2jags_2021] to fit the Bayesian model. We ran 15,000 iterations of 4 chain, discarded the first 10,000 interations, and used a thinning rate of 2. We assessed convergence for a small random subset of example trajectories using a threshold for the Gelman-Rubin diagnostic [@gelman_inference_1992] of r-hat < 1.1, and visually inspected posterior autocorrelation and traceplots produced by the _mcmcplots_ R package [@curtis_mcmcplots_2018].

@@

Following Dunham et al. [-@dunham_strategic_2023] we also calculate the expected value of sample information (EVSI) and the expected value of perfect information (EVPI) [@runge_which_2011;@yokota_value_2004], assigning a value of 1 to correct status assessments, and a value of 0 to incorrect assessments ($\nu_{\alpha,\psi}=1 \text{ when } \alpha=\psi, 0 \text{ otherwise}$). The expected value with perfect knowledge is $EV_{certainty}=1$. The difference between this and the expected score using knowledge of disturbance the disturbance scenario $\tau$ (Fig \@ref(fig:figDisturbance)) and priors from the national model gives EVPI, a measure of the maximum possible benefit that can be gained from local sampling:
\begin{equation}
EVPI_{\tau} = 1-max_{\alpha}\sum_\psi[P(\psi|\tau)\nu_{\alpha,\psi}],
  (\#eq:EVPI)
\end{equation}
where $P(\psi|\tau)$ is the prior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the prior. EVSI measures expected improvement from a particular sampling strategy as the difference between the expected value with and without additional sample information: 
\begin{equation}
EVSI_{\tau} = E_x[max_{\alpha}\sum_\psi[P(\psi|x,\tau)\nu_{\alpha,\psi}]] - (1-EVPI_{\tau}),
  (\#eq:EVSI)
\end{equation}
where $x$ is the sample information from a particular example trajectory. Simulated observations from 500 example trajectories for each disturbance and monitoring scenario $\tau$ (Figure \@ref(fig:figDisturbance)) represent the distribution of possible values of $x$. $P(\psi|x,\tau)$ is the posterior probability that the true state of the population is $\psi$, and each population is assigned a status $\alpha$ informed by the sample information.

@@

Analysis was implemented using the R language [@r_core_team_r_2023] and a variety of _tidyverse} packages [@wickham_welcome_2019]. To make the work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios, borrowing and building on code from an app developed by Eacker et al. [-@eacker_web-based_2019] where appropriate. Our app allows users to provide existing local data, or to make projections informed by simulated data as we have done in this analysis (similar to Figure
\@ref(fig:examplePredictions)). The app allows advanced users to adjust many model parameters (supplemental Table A1). Default parameter values are set to allow novice users to get reasonable results quickly. Most of the code needed to run the app is in an R package, _caribouMetrics_, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of other contexts [e.g. @dyson_existing_2022; @stewart_climate-informed_2023]. The _caribouMetrics_ package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>. Find documentation and vignettes on the package website
<https://landscitech.github.io/caribouMetrics/>. The Shiny user interface app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>. Code required to run analyses and reproduce figures in this paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.

@@

Analysis of simulated example demographic trajectories confirms that our Bayesian model can effectively combine local demographic data with priors from national demographic-disturbance relationships to reduce uncertainty in boreal caribou demographic projections (Figs \@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected, uncertainty about the population growth rate decreases as sampling effort in each year (i.e. number of collars) or the number of years of monitoring (Fig \@ref(fig:figDiffs)) increase. Uncertainty is higher when the forecast horizon is long (compare 5 and 20 year projections in Fig \@ref(fig:figDiffs)). Uncertainty is also higher in our high anthropogenic disturbance scenarios (Fig \@ref(fig:figDiffs)) because stochastic variation in population growth rates increase as populations become very small (as in top right panel of Fig \@ref(fig:examplePredictions)). These results confirm the Bayesian population model is working as it should.

Our analysis becomes more interesting and informative when we examine the probability of correctly assessing population status (Fig \@ref(fig:figStatusErrors)) and the expected value of information (Figs \@ref(fig:figEVSI) and \@ref(fig:figEVPI)). When anthropogenic disturbance is high (right panels of Figs \@ref(fig:figStatusErrors), \@ref(fig:figEVSI), and \@ref(fig:figEVPI)), projections of population status informed only by priors from the national demographic-disturbance relationships are accurate, and adding additional local monitoring data does not improve accuracy. Interestingly, when the projection period is long and anthropogenic disturbance is high, we see a negative EVSI (lower right panel of Fig \@ref(fig:figEVSI)), and we do not see an associated decrease in the probability of correctly assessing population status (lower right panel of Fig \@ref(fig:figStatusErrors)). This means that the long-term projections informed by local monitoring data are on average less precise but not less accurate, which can happen when there is very small prior probability that population is viable (orange band in the top right panel of Fig \@ref(fig:examplePredictions) falls below 1). At lower levels of disturbance, local monitoring can reduce uncertainty, but there are diminishing returns as the number of collars deployed or the number of years of monitoring increases (Figs \@ref(fig:figStatusErrors) and \@ref(fig:figEVSI)). The shapes of the relationships between monitoring effort and the probability of correct status assessment depend on how anthropogenic disturbance is changing over time, but there are diminishing benefits of increased monitoring regardless of whether disturbance is increasing or decreasing (supplement D). Adding unknown bias increases uncertainty, but these increases are not large for the set of parameters we investigate (supplement E).
